### 2020-04-14 report/draft.tex
Report draft

\documentclass{article}

\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{xspace}
\usepackage{color}
\usepackage{xcolor}
\usepackage{upquote}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{wrapfig}
\usepackage{syntax}

\usepackage{tikz}
\usetikzlibrary{arrows,automata,shapes.misc,shapes.geometric,positioning}

% For comments
\newcommand{\eat}[1]{}
\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\todo}[1]{\hl{#1}\xspace}

%% Bibliography style
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\begin{document}

\title{CIS 673 Project Report}
\author{Gautam Mohan and Konstantinos Kallas}

\maketitle

\section{Introduction}
\label{intro}

\section{Servois}

\section{Conflict Free Replicated Data Types}

\TODO{Fill in with some background information on CRDTs}~\cite{shapiro2011conflict}

\section{Methodology}

\section{Experience Report}

%% Bibliography
\bibliography{./draft}
\bibliographystyle{plain}

\end{document}

### 2020-04-15 report/draft.tex
report todos

@@
 \section{Introduction}
-\label{intro}
+
+\TODO{Motivate CRDTs and why it makes sense to verify them
+  automatically.}
 
 \section{Servois}
+
+\TODO{Describe the technique briefly} \TODO{Describe the tool
+  interface. What exactly one has to encode and why}
 
@@
 
+\subsection{Last Writer Wins Register}
+
+We first encoded the simple last writer wins register from
+\cite{shapiro2011comprehensive}. Its specification is shown in
+\Cref{fig:lww-def}. \TODO{Describe the LWW Register.}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\textwidth]{lww-definition}
+    \caption{Last Writer Wins Register}
+    \label{fig:lww-def}
+\end{figure}
+
+\TODO{Describe how we implemented it, and why we need to have a set of seen Timestamps to enforce uniqueness.}
+
+\subsection{Grow-only Set and Two-Phase Set}
+
+We have encoded the Grow-only Set (GSet) and Two-Phase Set (2PSet)
+from \cite{baquero2017pure}. GSets only support adding items, and they
+can be useful when logging events where logging is unimportant
+(e.g. when storing IPs of clients visiting a web-site). 2PSets extend
+GSets by supporting removals, but items that have been removed can
+never be added again. Their definitions are shown in
+\Cref{fig:get-def,fig:2pset-def}. The state of a GSet is just an
+originally empty set of values of type $V$. It's only update function
+is \texttt{add} which does nothing in the \texttt{prepare} stage, and
+then updates the state by inserting the input element $v$ in the set
+$s$. The two supported queries are asking for the elements and the
+size of the state.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.5\textwidth]{grow-only-set-definition}
+    \caption{Grow-only Set}
+    \label{fig:gset-def}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.5\textwidth]{2pset-definition}
+    \caption{Two-Phase Set}
+    \label{fig:2pset-def}
+\end{figure}
+
+\subsection{Graph CRDTs}
+
+Say some things about the graph in section 5 of
+\cite{shapiro2011conflict} and specification 16 from
+\cite{shapiro2011comprehensive}.
+
+\TODO{Describe where we got stuck and why.}
+
 %% Bibliography

### 2020-04-15 test/runall.py
Faulty lwwreg

@@ -36,6 +36,10 @@ FULLRESULTS="fullresults.tex"
 
 # datastructures=['VectorClock']
 datastructures=['CRDTGraph']
+datastructures=['CRDTGraphAlt']
+datastructures=['2PGraph']
+datastructures=['2PSet']
+datastructures=['LWWReg']
 # datastructures=['CRDTOpCounter']
 #datastructures=['Hashtable']
 #datastructures=['Counter']
@@ -49,7 +53,11 @@ labels={'Counter':'counter',
         'Stack':'stack',
         'VectorClock':'vectorclock',
         'CRDTOpCounter':'crdtopcounter',
-        'CRDTGraph':'crdtgraph'}
+        'CRDTGraph':'crdtgraph',
+        'CRDTGraphAlt':'crdtgraphalt',
+        '2PGraph':'2pgraph',
+        '2PSet':'2pset',
+        'LWWReg':'lwwreg'}
 forceruns=[]
 bestopt = '--poke'
 

### 2020-04-15 report/draft.tex
Start writing the LWWReg in the report

@@
 
+In this section we describe our experience from proving that several
+operation-based CRDTs commute. We briefly describe the semantics of
+each CRDT and then explain how we encoded them in the Servois
+interface.
+
 \subsection{Last Writer Wins Register}
@@
 \cite{shapiro2011comprehensive}. Its specification is shown in
-\Cref{fig:lww-def}. \TODO{Describe the LWW Register.}
+\Cref{fig:lww-def}. A last writer wins register is a simple register
+data type with one value that can be updated and read. It creates a
+total order of assignments by associating a unique timestamp with each
+update. By always keeping the value with the latest timestamp its
+updates commute with each other.
 
@@
 \end{figure}
+
+In order to prove that the LWW Register is Strongly Eventually
+Consistent, we need to prove that all different concurrent effects
+(indicated as ``downstream'' in \cref{fig:lww-def}) commute. Since
+only the value $x$ of the register can be returned using queries, we
+consider two states to be equivalent if they have the same value $x$.
+
+We started by encoding the state as is (a value $x$ of any type $X$
+and a timestamp $t$ of type $Int$) and by encoding the $assign$ effect
+by only requiring as a precondition that $t' \neq t$, in order to
+enforce timestamp uniqueness. \TODO{Finish this}
+
+%% Servois quickly retuned the following
+%% counterexample:
+
+%% \[
+%% x_1 \neq x_2 \wedge t_1 = t_2 \wedge t
+%% \]
 

### 2020-04-15 report/draft.tex
Write more on the last writer wins register

@@
 by only requiring as a precondition that $t' \neq t$, in order to
-enforce timestamp uniqueness. \TODO{Finish this}
+enforce timestamp uniqueness. Servois quickly retuned the following
+counterexample:
 
-%% Servois quickly retuned the following
-%% counterexample:
+\[
+t_1' = t \vee t_2' = t
+\]
 
-%% \[
-%% x_1 \neq x_2 \wedge t_1 = t_2 \wedge t
-%% \]
+\TODO{Gautam: Do you understand why exactly does this fail? It is in
+  line 36,37 of lwwreg.yml} Servois found that if one of the effect
+timestamps $t_1', t_2'$ is equal to the initial timestamp $t$, then
+executing them in different orders will lead to one of them not
+satisfying the precondition once, but satisfying it in the other
+run. However, this behaviour is disallowed by the LWW specification
+since all new timestamps $t'$ are unique.
+
+In order to deal with that, we extended the state to keep a set of all
+seen timestamps, and we modified the precondition to require that
+\TODO{Finish this.}
 

### 2020-04-15 report/draft.tex
Write part of methodology

@@
 
+\TODO{Say something about both State and Op-based CRDTs.}
+
+\TODO{State correctness conditions Op-based only}
+
+\TODO{Say why we only need to prove commut of effects}
+
 \section{Methodology}
+
+As we mentioned above, in this project we focus on operation based
+CRDTs, since proving commutativity of the effects of their supported
+operations implies that they are Strongly Eventually Consistent. In
+this section, we describe our methodology and informally argue about
+its soundness.
+
+In order to encode a data type in Servois, we need to specify (i) its
+state, (ii) an equivalence relation on states, (iii) each supported
+method (their arguments, preconditions, and postconditions), and (iv)
+a set of predicates that can be used by Servois to iteratively refine
+the state space. On the other hand, an op-based CRDT specification
+contains the following components: (i) a state definition, (ii) a set
+of query methods, and (iii) a set of update methods, each of which has
+a prepare and an effect method.
+
+We start by encoding the state without any modifications. State
+equivalence is defined as a conjunction of the equivalence of all
+different state components. Query operations do not be to be encoded
+as methods in Servois as they never update the state.
+%% \TODO{Check in the CRDT paper if equivalence on states is defined
+%%   based on the query methods or based on a complete equivalence of
+%%   the states. Accordingly complete this paragraph describing how to
+%%   define equivalence.}
+
+We now describe how to encode the update methods. Each update method
+has two parts, a \texttt{prepare} and an \texttt{effect}. In Servois
+we only encode effects as methods, as we care about how concurrent
+updates (that happened in different replicas) can affect the state of
+another replica. Intuitively, we want to show that the effects $e_1$,
+$e_2$ of two concurrent updates $u_1$, $u_2$ that happened in replicas
+$r_1$ and $r_2$ will commute in a third replica $r_3$. In order to
+correctly do that, we need to derive preconditions for all the effect
+methods by using their prepare methods. More precisely, the
+precondition on the arguments of an effect $e$ of an update $u$ that
+happened in a replica with state $s'$ must be equivalent to the
+postcondition of $u$'s prepare method after it has been applied
+$s'$. \TODO{Is that valid?}
+
+\TODO{Figure out what we have to do with effect preconditions. Maybe
+  we need to encode them with ite to make sure that if they fail, it
+  means that an update is not enabled when it should be?}
+
+Finally, since Servois is sound, i.e. it never concludes that two
+methods commute if they don't, we decided to leave the predicates
+empty and gradually extend them if we find that Servois is unable to
+give an answer \TODO{Check the validity of the above statement by
+  reading the Servois paper}.
+
+\TODO{Briefly mention the threats to validity, the fact that we derive
+  preconditions based on prepare methods but we don't check them, and
+  the fact that we manually encode the postconditions of each method.}
 
@@
 from \cite{baquero2017pure}. GSets only support adding items, and they
-can be useful when logging events where logging is unimportant
+can be useful when logging events where order is unimportant
 (e.g. when storing IPs of clients visiting a web-site). 2PSets extend
@@
 never be added again. Their definitions are shown in
-\Cref{fig:get-def,fig:2pset-def}. The state of a GSet is just an
-originally empty set of values of type $V$. It's only update function
-is \texttt{add} which does nothing in the \texttt{prepare} stage, and
-then updates the state by inserting the input element $v$ in the set
-$s$. The two supported queries are asking for the elements and the
-size of the state.
+\Cref{fig:gset-def,fig:2pset-def}.
 
@@
 
+The state of a GSet is just an originally empty set of values of type
+$V$. It's only update function is \texttt{add} which does nothing in
+the \texttt{prepare} stage, and then updates the state by inserting
+the input element $v$ in the set $s$. The two supported queries are
+asking for the elements and the size of the state. The specification
+of 2PSet can be seen as a composition of 2 GSets, one for additions
+and one for removals. The removals set prohibits items from being
+readded to the additions set. 2PSet supports two updating functions,
+\texttt{add $v$} and \texttt{remove $v$}. Addition inserts the element
+in the addition set only if it doesn't not occur in the removal set,
+and removal removes an element from the addition set and adds it in
+the removal set. The intuition behind the commutativity of 2PSet is
+that removals don't need to happen after an addition to remove the
+item, so reversing the order of an addition and a removal of the same
+value results in the item being removed.
+
 \begin{figure}[h]
@@
 \end{figure}
+
+Encoding the two specifications in Servois is mostly straightforward
+since all the \texttt{prepare} methods are identity functions without
+preconditions. We just encode the result of their effects as
+postconditions. Here is an excerpt of the specification of 2PSet,
+specifically the state definition and the effect of the \texttt{add}.
+
+\begin{verbatim}
+  state:
+    - name: Add
+      type: (Set V)
+    - name: Remove
+      type: (Set V)
+
+  methods:
+    - name: add
+      args:
+        - name: v
+          type: V
+      return:
+        - name: result
+          type: Bool
+      requires: |
+        true
+      ensures: |
+        (and result
+             (= Add_new (union (setminus (singleton v) Remove) Add))
+             (= Remove_new Remove))
+\end{verbatim}
 

### 2020-04-15 report/draft.tex
Sections 3 and 4

@@
 
-\title{CIS 673 Project Report}
+\title{Commutativity of operation-based CRDTs \\ CIS 673 Project Report}
 \author{Gautam Mohan and Konstantinos Kallas}
@@
 
-\section{Conflict Free Replicated Data Types}
+\section{Conflict-Free Replicated Data Types}
 
-\TODO{Fill in with some background information on CRDTs}~\cite{shapiro2011conflict}
+Conflict-free replicated data types are objects that can be
+implemented on top of eventually consistent datastores guaranteeing
+strong convergence without any need for conflict resolution or
+roll-back. Conflict-freedom ensures safety and liveness despite any
+number of failures. This section contains necessary background
+information for CRDTs that is adapted from~\cite{shapiro2011conflict}.
 
-\TODO{Say something about both State and Op-based CRDTs.}
+The standard correctness condition for conflict-free replicated data
+types is strong eventual consistency (SEC), which is slightly stronger
+than eventual consistency. Its main difference is strong convergence,
+namely that replicas that have delivered the same updates have
+equivalent states. Objects that are SEC can be implemented on top of
+replicated data stores that do not synchronize, allowing them to be
+always available for both reads and writes independently of network
+conditions.
 
-\TODO{State correctness conditions Op-based only}
+Conflict-free replicated data types can be classified in two types,
+the state-based ones and the operation-based ones. The intuitive
+difference is that in state-based CRDTs replicas merge their states
+periodically, while in operation-based CRDTs all replicas broadcast
+all of their updates to all other replicas. Both types of CRDTs can be
+proven to be SEC given some conditions on their specifications
+hold. In this project we focus on operation-based CRDTs, since their
+correctness, i.e. satisfying strong eventual consistency, is implied
+by their commutativity of their operations.
 
-\TODO{Say why we only need to prove commut of effects}
+\subsection{Operation-based CRDTs}
+
+An op-based object can be specified by the following tuple $(S, s^0,
+Q, U)$, where $S$ is the state domain, $s^0$ is the initial state, $Q$
+is a set of query methods, and $U$ are update methods. Update methods
+are split into a pair $(t,u)$, where $t$ is a side-effect-free
+\texttt{prepare} method and $u$ is an \texttt{effect} method (the
+arguments of which may differ. Both \texttt{prepare} and
+\texttt{effect} methods have enabling preconditions. The
+\texttt{prepare} method is executed at a single replica where the
+operation is invoked, generating the arguments for the \texttt{effect}
+without modifying the state. The \texttt{effect} is executed
+immediately after it in that replica, and it is then broadcast to all
+other replicas for execution.
+
+An important assumption is that the underlying broadcast communication
+protocol delivers messages preserving causal order. This means that if
+a \texttt{prepare} $t_j$ is executed on a replica after it has
+received an effect $e_i$, then all other replicas will receive $e_j$
+after they receive $e_i$. Concurrent updates (i.e. ones that happened
+in different source replicas can be delivered in any order.
+
+Given these assumptions, they prove in
+Theorem~2.2~\cite{shapiro2011conflict} that the commutativity of all
+concurrent effect operations of an op-based CRDT is a sufficient
+condition for convergence. Note that they make one more assumption,
+namely that causally-ordered broadcast is sufficient to ensure that an
+effect is enabled in all other replicas if it was enabled in the
+initial one.
+
+%% Effect precondition may require that an effect is delayed, so in order
+%% to prove convergence, one would need to prove liveness, that
+%% eventually an effects precondition is satisfied. They only focus on
+%% effect preconditions that are satisfied by the delivery order in that
+%% paper, and that is what we focus on too.
 
@@
 has two parts, a \texttt{prepare} and an \texttt{effect}. In Servois
-we only encode effects as methods, as we care about how concurrent
-updates (that happened in different replicas) can affect the state of
-another replica. Intuitively, we want to show that the effects $e_1$,
-$e_2$ of two concurrent updates $u_1$, $u_2$ that happened in replicas
-$r_1$ and $r_2$ will commute in a third replica $r_3$. In order to
-correctly do that, we need to derive preconditions for all the effect
-methods by using their prepare methods. More precisely, the
-precondition on the arguments of an effect $e$ of an update $u$ that
-happened in a replica with state $s'$ must be equivalent to the
-postcondition of $u$'s prepare method after it has been applied
-$s'$. \TODO{Is that valid?}
+we only encode effects as methods, as we are only interested in the
+commutativity of concurrent updates.
+%% \TODO{Briefly mention that we don't care about commutativity of
+%%   prepare-effect with effect because of the underlying causal
+%%   delivery protocol.}
+Intuitively, we want to show that the effects $e_1$, $e_2$ of two
+concurrent updates $u_1$, $u_2$ that happened in replicas $r_1$ and
+$r_2$ will commute in a third replica $r_3$. In order to correctly do
+that, we need to derive preconditions for all the effect methods by
+using their prepare methods. More precisely, the precondition on the
+arguments of an effect $e$ of an update $u$ that happened in a replica
+with state $s'$ must be equivalent to the postcondition of $u$'s
+prepare method after it has been applied $s'$.
+%% \TODO{Is that valid?}
 
-\TODO{Figure out what we have to do with effect preconditions. Maybe
-  we need to encode them with ite to make sure that if they fail, it
-  means that an update is not enabled when it should be?}
+\TODO{(Maybe -- If we have time) Figure out what we have to do with
+  effect preconditions. Maybe we need to encode them with ite to make
+  sure that if they fail, it means that an update is not enabled when
+  it should be?}
 
@@
 
-\TODO{Briefly mention the threats to validity, the fact that we derive
-  preconditions based on prepare methods but we don't check them, and
-  the fact that we manually encode the postconditions of each method.}
+\paragraph{Threats to validity:}
+
+We would like to briefly mention two threats to the validity of our
+methodology. First, both \texttt{effect} pre and postconditions are
+derived from the specification manually, introducing a possible point
+of error. Second, we do not check whether the delivery preconditions
+of \texttt{effect} methods are indeed satisfied by the causal delivery
+order (similarly to~\cite{shapiro2011conflict}).
+
+%% \TODO{Briefly mention the threats to validity, the fact that we derive
+%%   preconditions based on prepare methods but we don't check them, and
+%%   the fact that we manually encode the postconditions of each
+%%   method. Also that we don't check whether delivery preconditions (of
+%%   effects) are indeed satisfied by the causal delivery order.}
 
@@
     \centering
-    \includegraphics[width=0.5\textwidth]{grow-only-set-definition}
+    \includegraphics[width=0.4\textwidth]{grow-only-set-definition}
     \caption{Grow-only Set}

### 2020-04-17 report/draft.tex
Sections 1,2,5.3

@@
 
-\TODO{Motivate CRDTs and why it makes sense to verify them
-  automatically.}
+Modern day distributed programming presents many challenges. One
+key issue that has been well-studied is the idea of maintaining consistency
+between replicated instances of data. There are many algorithms and
+protocols that present solutions with differing tradeoffs. One major
+source of complexity that occurs across various solutions is the idea of
+resolving conflicts. If two replicated instances of data receive conflicting
+updates separately, how should we specify the merge of these two versions?
+
+Conflict-Free Replicated Data Types (CRDTs) are a recent solution that attempts
+to sidestep the issue of merge conflicts. These data structures have semantics
+for merging built in to their specification, and ensure that updates made in any
+order at any replicas will always commute, and therefore never be in conflict.
+However, one downside of these guarantees is the limited sorts of data
+structures that have been currently proven to be CRDTs. Furthermore, CRDT
+correctness proofs must currently be done manually.
+
+The goal of our project is to reduce the burden of manually verifying that
+methods of CRDTs commute (a necessary condition to them being conflict-free). We
+employ a tool called Servois that is able to automatically generate
+commutativity conditions given pre- and postconditions of an abstract data type.
+By helping to automate the verification process of CRDTs, we hope that more data
+structures can be proven to be conflict-free, which would allow programmers
+richer abstractions they could use to write programs.
 
@@
 
-\TODO{Describe the technique briefly} \TODO{Describe the tool
-  interface. What exactly one has to encode and why}
+Servois \cite{Bansal2018Servois} is a tool that automatically computes the
+commutativity (and non-commutativity) conditions of abstract data types (ADTs). Unlike
+previous tools that relied on random sampling or blackbox testing, Servois is
+able to generate conservative approximations of commutativity conditions that
+are guaranteed to be sound, even if they are partial. Servois is
+\textit{relatively complete}: if it terminates, the commutativity conditions are
+guaranteed to be complete but termination is not guaranteed.
+
+Servois' algorithm can be thought of as similar to counterexample-guided
+abstraction refinement. At a high-level, it starts with a coarse partitioning of
+the state space and recursively divides it based on increasingly stronger
+predicates. Given a partition of the state space corresponding to a certain
+predicate, Servois first checks if the ADT is commutative or non-commutative
+given that predicate as a precondition on inputs. If it is, no more work is
+performed and the predicate is added to the commutativity conditions. If there
+are inputs satisfying the predicate that are both commutative and
+non-commutative, Servois uses them as counterexamples and partitions the space
+recursively. It stops only when the entire state space has been partitioned
+completely into commutative or non-commutative sections, or if the user manually
+halts the algorithm, in which case Servois returns an overapproximation of the
+commutativity conditions.
+
+Servois is an ideal tool to use in this project to alleviate the burden of
+performing manual proofs on whether or not methods of a CRDT commute. The only
+input it requires is the methods of the datatype whose correctness conditions
+are specified as pre- and post-conditions in decidable SMT theories. Since CRDTs
+are meant to be specified the same as ordinary ADTs, it is straightforward to
+encode them in this format. For our project we have specified the correctness
+pre- and postconditions manually, but it would be interesting future work to see
+if these invariants could be extracted automatically based on the code.
 
@@
 
-Say some things about the graph in section 5 of
-\cite{shapiro2011conflict} and specification 16 from
-\cite{shapiro2011comprehensive}.
+Our final attempt to encode a CRDT was the directed graph specified in Section 5
+of \cite{shapiro2011conflict}. It allows for conflict-free queries, additions,
+and deletions of nodes and edges. The specification for the directed graph CRDT
+can be seen in \Cref{fig:graph-def}. At a high-level, this graph allows for
+simultaneous additions and deletions of vertices and edges across replicas.
+These updates are reflected as sets representing the change made by each
+operation, and periodically these deltas are sent as updates to all other
+replicas. The mechanism by which we are able to handle duplicate
+insertions/references is by associating every vertex and edge with a unique tag.
 
-\TODO{Describe where we got stuck and why.}
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=1\textwidth]{directed-graph-def}
+    \caption{Op-based Directed Graph Specification}
+    \label{fig:graph-def}
+\end{figure}
+
+We were able to come up with a straightforward translation of the spec using
+CVC4's theory of sets and arrays. However, there were two main issues that
+prevented Servois from being able to find commutativity conditions. The
+\textit{removeVertex} and \textit{removeArc} (edge) preconditions have an
+$\exists$ condition, and CVC4 would immediately return ``unknown'' when queried.
+Though we do not know exactly why this occurred, we suspect it is because of the
+nested quantification that occurs when introducing an $\exists$ clause in the
+specification. Even when we refactored the logic to use a $\forall$, CVC4 still
+could not solve the query.
+
+Many of the more complex CRDT specifications contained specifications that
+involved quantifiers. It would be interesting to see if these could be
+refactored to be quantifier-free. It is important to note that all the example
+CRDTs in the Servois paper had quantifier-free encodings. Furthermore, their
+method was careful to come up with alternative formulations that avoided
+``$\forall\exists$'' quantifier alternation, which they noted would make queries
+undecidable \cite{bansal2018servois}.
 

### 2020-04-17 report/draft.tex
This is correct. Servois is sound and "relatively complete".

@@
 empty and gradually extend them if we find that Servois is unable to
-give an answer \TODO{Check the validity of the above statement by
-  reading the Servois paper}.
+give an answer.
 

### 2020-04-18 report/draft.tex
Change a capital letter citation

@@
 
-Servois \cite{Bansal2018Servois} is a tool that automatically computes the
+Servois \cite{bansal2018servois} is a tool that automatically computes the
 commutativity (and non-commutativity) conditions of abstract data types (ADTs). Unlike

### 2020-04-18 report/draft.tex
Finalize report draft

@@
 
-\TODO{(Maybe -- If we have time) Figure out what we have to do with
-  effect preconditions. Maybe we need to encode them with ite to make
-  sure that if they fail, it means that an update is not enabled when
-  it should be?}
+%% \TODO{(Maybe -- If we have time) Figure out what we have to do with
+%%   effect preconditions. Maybe we need to encode them with ite to make
+%%   sure that if they fail, it means that an update is not enabled when
+%%   it should be?}
 
@@
 Consistent, we need to prove that all different concurrent effects
-(indicated as ``downstream'' in \cref{fig:lww-def}) commute. Since
-only the value $x$ of the register can be returned using queries, we
-consider two states to be equivalent if they have the same value $x$.
+(indicated as ``downstream'' in \cref{fig:lww-def}) commute. We
+consider the Since only the value $x$ of the register can be returned
+using queries, we consider two states to be equivalent if their values
+$x$ and timestamps $t$ are equal.
 
-We started by encoding the state as is (a value $x$ of any type $X$
-and a timestamp $t$ of type $Int$) and by encoding the $assign$ effect
-by only requiring as a precondition that $t' \neq t$, in order to
-enforce timestamp uniqueness. Servois quickly retuned the following
-counterexample:
+We started by encoding the state as a value $x$ of any type $X$ and a
+timestamp $t$ of type $Int$. In order to encode the \texttt{effect} of
+the $assign$ method we need to ensure that the timestamp $t'$ is
+unique. In order to achieve that, we need to extend the state with an
+additional component $SeenTimestamps$ that contains the set of all
+timestamps that have been encountered so far. This auxiliary component
+can be used to encode uniqueness as the \texttt{effect}-method
+precondition:
 
 \[
-t_1' = t \vee t_2' = t
+t' \not \in SeenTimestamps \wedge t \in SeenTimestamps
 \]
 
-\TODO{Gautam: Do you understand why exactly does this fail? It is in
-  line 36,37 of lwwreg.yml} Servois found that if one of the effect
-timestamps $t_1', t_2'$ is equal to the initial timestamp $t$, then
-executing them in different orders will lead to one of them not
-satisfying the precondition once, but satisfying it in the other
-run. However, this behaviour is disallowed by the LWW specification
-since all new timestamps $t'$ are unique.
-
-In order to deal with that, we extended the state to keep a set of all
-seen timestamps, and we modified the precondition to require that
-\TODO{Finish this.}
-
-\TODO{Describe how we implemented it, and why we need to have a set of seen Timestamps to enforce uniqueness.}
+\noindent
+This ensures that all new timestamps cannot be in the set of seen
+timestamps and that the current timestamp is in the set (which is an
+invariant). We then specified the effect method postconditions as seen
+in its specification. Servois successfully verified that the effect
+method commutes with itself.
 
